### 2016-06-30 stitched.Rmd
ENH:  More revisions.

header-includes:
   - \usepackage{booktabs}
   - \usepackage[final]{changes}
   - \usepackage[font={small},labelfont=bf,labelsep=colon]{caption}
   - \linespread{2}
   - \usepackage[compact]{titlesec}
   - \usepackage{enumitem}
   - \usepackage{tikz}
   - \def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
   - \setlist{nolistsep}
   - \titlespacing{\section}{2pt}{*0}{*0}
   - \titlespacing{\subsection}{2pt}{*0}{*0}
   - \titlespacing{\subsubsection}{2pt}{*0}{*0}
   - \setlength{\parskip}{3pt}
   - \setremarkmarkup{(#2)}
   - \definechangesauthor[name={Nick Tustison}, color=red]{nt}
   - \definechangesauthor[name={James Stone}, color=magenta]{js}
   - \definechangesauthor[name={Lisa Wilde}, color=cyan]{lw}
   - \definechangesauthor[name={Andy Mayer}, color=green]{am}
   - \definechangesauthor[name={Harvey Levin}, color=blue]{hl}
   - \definechangesauthor[name={Brian Taylor}, color=purple]{bt}
   - \definechangesauthor[name={David Tate}, color=orange]{dt}
bibliography: references.bib
csl: council-of-science-editors.csl
fontsize: 11pt
mainfont: Georgia
geometry: margin=1.0in
---

@@

White matter hyperintensities (WMHs) are foci of abnormal signal
intensity in white matter regions seen with magnetic resonance imaging
(MRI). WMHs are associated with normal aging and have shown prognostic
value in neurological conditions such as traumatic brain injury (TBI).
The impracticality of manually quantifying these lesions limits their
clinical utility and motivates the utilization of machine learning
techniques for automated segmentation workflows.  Herein, we develop a
concatenated random forest framework with image features for
segmenting WMHs in a TBI cohort. The framework is built upon the
Advanced Normalization Tools (ANTs)  and ANTsR toolkits.  MR
(3D FLAIR, T2-, and T1-weighted) images from 24 service members and
veterans scanned in the Chronic Effects of Neurotrauma Consortium’s
(CENC) observational study were acquired. Manual annotations were
employed for both training and evaluation using a leave-one-out
strategy.  Lesion load and overlap evaluative comparisons are
complimented by feature rankings which showcase the utility of the
concatenated approach.  Our findings suggest supervised learning methods
may be applied to quantify WMHs on routine brain imaging. Paired with
correlative outcome data, supervised learning methods may allow for
identification of imaging features predictive of diagnosis and prognosis
in individual TBI patients.

\clearpage
<!--
A couple notes in response to comments:

[David Tate, Comment 15]:  Stupid question, but why not show the same subject and slice for both for a better comparison?

NT:  I see what he is saying and probably agree.

[David Tate, Comment 28]:  Maybe this could go in a Figure as a flow chart??

NT:  Flow charts are great and I included a couple as Figure 5 in my tumor paper.  However, they take a long time to make correctly and I do not want to commit to producing one unless it is absolutely necessary.

[Andy Mayer, Comment 30]: It will be difficult to see this font for labeling on Y axis at current size/resolution. You may also consider alerting your reader directly that the magnitudes of the scales are very different for stage 1 versus stage 2 in the legend or scale them to be similar.

NT:  Comparing these two plots in Figure 4 in terms of their absolute MeanDecreaseAccuracy
is meaningless so coordinating the scales is not important.

[Andy Mayer, Comment 34]:  A reviewer is likely to ask for some sort of reliability rating on the labeling of the lesions.

NT:  I agree with this but unless somebody else is willing to manually segment the WMHs,
then there is not really much we can do other than point to other papers which have looked
at reliability.

[David Tate, Comment 37]:  Seems like we are missing a picture.  That is the football player in me, but it seems like we need a picture that shows the manual compared to the automated at each stage if possible.

NT:  We could do this but I do not think that such images in publications are that meaningful as they typically result from heavy selection bias.  I did do something similar
for my brain segmentation paper where I showed multiple slices (skipped every 10, or something like that) for each subject in the challenge cohort but tumors are obviously much bigger and can be displayed in this manner.

[David Tate, Comment 39]:  I would would move these comments down to a limitation like paragraph, otherwise it minimizes the cool findings.

NT:  I have no opinion on this.  Do what you think is best.

[Andy Mayer, Comment 40]:  I do not know if you can include any of this data (even one subject). It would be great to demonstrate that your algorithm worked on data collected on another scanner since the intensities are likely to be very different.

NT:  As I mentioned to you previously, I think this is a great idea and I would be happy to include it but how do we showcase this unless somebody is willing to annotate data from these other sites.  We could include a picture of a single subject but how persuasive that is will be reviewer-dependent.  Personally, I would not be persuaded as the showing the results of a single subject is very prone to selection bias.

@@

\textcolor{blue}{Figure 1 provides a graphical overview of the proposed workflow.  The major components
include offline generation of symmetric multimodal templates, the creation of
feature images from the training data which are then employed for statistical
prediction using a concatenated random forest framework.  This framework involves
the use of two random forest models where the outcome (i.e., the tissue
probability estimates) of the first RF model application,
which we denote as ``Stage1 '', is used as input (along with the original set of
feature images) to a refinement RF model segmentation which we denote as ``Stage 2''.
Once these offline steps
are performed, a new, unsegmented subject can then be processed using the proposed pipeline.}

@@

<!--
![Canonical views of the mutlivariate, bilaterally symmetric template constructed
from the MMRR data set [@landman2011] (only shown are the FLAIR, T1, and T2 modalities---
the components relevant for this work).  Template construction is detailed in
[@Tustison:2015aa].  These images are important for specific asymmetry-based
features.](Figures/MMRR.png)

@@

\textcolor{blue}{In Figure 8 we provide the segmentation evaluations derived from the leave-one-out evaluation
of the previously described TBI data.  These performance measures include sensitivity,
positive predictive value, $F_1$ score, and relative volume difference.  The three lesion
size ranges over which these measures are computed to illustrate the variation
in performance with lesion size.  Smaller lesions ($< 12$ voxels) are more difficult
to identify which is why the sensitivity is widely varied over the cohort vs. the largest
set of voxels ($> 28$ voxels).   The first three measures are based on the identification of
entire lesions.  The relative volume difference provides a direct assessment of how accurate
the volumetric estimate is when comparing the manually identified lesions versus the
automatically predicted lesions.}

@@

In Figure 8 are the segmentation comparisons derived from manual segmentations of the same
data.  Despite the large variability characteristic with manual labelings in related fields
[@Grimaud:1996aa;@styner2008;@Garcia-Lorenzo:2013aa], such labelings are characteristic of
current clinical practices and the methodology proposed herein is readily adapted to refinements
in training data.

On the left of Figure 8 are the improvement in Dice values [@tustison2009],
over all white matter hyperintensities when comparing the segmentations between the two stages
where the sum is taken over all individually labeled manual, $T_r$, and automated, $S_r$, lesions and $\cap$
represents the intersection between the manual/automated lesion pair.  Performing
the second round of supervised learning improves these Dice values.  One can also note from the
right side of Figure 7 that the total lesion load volume illustrates a few subjects that are
severe outliers in terms of the number of false positives.  The second round helps to
correct this issue.

@@

![Evaluation measures for the leave-one-out protocol of the described protocol in the Methods
section:  (a) sensitivity, (b) positive predictive value, (c) $F_1$ score, and (d) relative
volume difference.   The three quantile ranges include the following lesion volume ranges:
$[3-12)$, $[12-28)$, and $[28-551]$ based on the manually-derived lesion volumes.  These
measures are computed for both Stages.  Overall improvement is seen as the
second Stage RF model is applied for all three quantile ranges.](Figures/evaluationResults.png)

@@

The resulting rankings for both Stages are given in Figures 5 and 6 where the values for the
separate stages are averaged over the entire corresponding model set.  In addition, we
track the variance for each feature over all models to illustrate the stability of
the chosen features during the evaluation.  This latter information is illustrated as
horizontal errors bars providing the $95^{th}$ percentile
Note that the reader can cross reference Table 1 for identifying corresponding feature types
and names.

<!--
One can also use these measurements as a type of sanity check.  For example, from the Stage
1  plot, one can see that the _MeanDecreaseAccuracy_ values for the location indices in the
anterior-posterior direction (i.e., _TemplateIndicesWarped1_) are greater than
those for either the inferior-superior (i.e., _TemplateIndicesWarped2_) or the
left-right (i.e., _TemplateIndicesWarped0_) directions in the space of the symmetric
template.
-->

Additionally, it is interesting to note some of the other top performing features for Stage 1.
The contralateral difference FLAIR image is  highly discriminative over the set of
evaluation random forest models (see Figure 7).  This accords with the known clinical relevance of
FLAIR images for identifying white matter hyperintensities and the fact that such
pathology does not typically manifest symmetrically in both hemispheres.  Interestingly, the
posterior maps for the deep gray matter are extremely important for accurate white matter
hyperintensity segmentation.  Perhaps the spatial specification of deep
gray matter aids in the removal of false positives.  Inspection of the bottom of the
plots demonstrates the lack of discriminating features associated with the
T1 image which is also well-known in the clinical literature.

As described earlier, for Stage 2, we used the output random forest voting maps from Stage
1 as both features themselves and as priors for input to a Bayesian-based segmentation with
an additional MRF spatial prior.  In Figure 6, the voting maps are labeled as
"_RFStage1VotingMaps_" where the final numeral is associated with the brain
parenchymal labeling given previously.  Similarly, the additional RF prior segmentation
feature probability maps are labeled as "_RFBrainSegmentationPosteriors_".
The Stage 2 feature importance plot follows similar
trends as that for Stage 1 with the T1 images not contributing much to the identification
of white matter hyperintensity voxels.  The initial voting maps from Stage 1 are extremely
important with the top 3 being the estimated locations of the 1) gray matter, 2) white matter, and
3) white matter hyperintensities.  Since these tissue type can be conflated based on intensity
alone it is intuitive that such features would be important.

@@

@@

# Figure Captions

__Figure 1:__ \textcolor{blue}{Workflow illustration for the proposed pipeline.  Processing of the multi-modal
input MRI for a single subject, using the multi-modal symmetric template, results in
the generation of the feature images.  These feature images are used as input to the
Stage 1 RF model producing the initial RF probability map estimates.  The Stage 1
voting maps, the original feature images, and the Stage 2 RF model result in the
final voting maps which includes the WMH probability estimate.  Note that the RF models
are constructed once from a set of training data which are processed using the
same feature-construction pipeline as the single-subject input MRI.}

__Figure 2:__  Canonical views of the mutlivariate, bilaterally symmetric template constructed
from the MMRR data set [@landman2011] (only shown are the FLAIR, T1, and T2 modalities---
the components relevant for this work).  Template construction is detailed in
[@Tustison:2015aa].  These images are important for asymmetry-based
features.

__Figure 3:__  Representation of Stage 1 feature images for subject 01C1019.  The
FLAIR, T1-, and T2-weighted images are rigidly pre-aligned
[@Avants:2014aa] to the space of the T1 image.  The three modality images are then preprocessed
(N4 bias correction [@Tustison:2010ac] and adaptive denoising [@Manjon:2010aa]) followed by
application of standard ANTs brain extraction and $n$-tissue segmentation protocols using
the MMRR symmetric template and corresponding priors [@Tustison:2014ab] applied to the T1 image.
The feature images are then generated for voxelwise input to the RF model which results in
the voting maps illustrated on the right.  This gives a probabilistic classification of
tissue type.  Not shown are the probability and voting images for the brain stem and
cerebellum.

__Figure 4:__  Sample FLAIR acquisition image slices showing both manual and random forest
segmentations for both stages obtained during the leave-one-out evaluation.  Manual
segmentations were performed by one of the authors and provided the ground
truth WMH labels for training the random forest models.

__Figure 5:__ Average _MeanDecreaseAccuracy_ plots generated from the creation of all 24 random
forest models for Stage 1 during the leave-one-out evaluation.  These plots
are useful in providing a quantitative assessment of the predictive importance of each feature.
Features are ranked in descending order of importance.
The horizontal error bars provide the $95^{th}$ percentile
and illustrate the
stability of the feature importance across the leave-one-out models.
At this initial stage only 31 feature images are used.

__Figure 6:__  Average _MeanDecreaseAccuracy_ plots generated from the creation of all 24 random
forest models for Stage 2 during the leave-one-out evaluation.  These plots
are useful in providing a quantitative assessment of the predictive importance of each feature.
Features are ranked in descending order of importance.
The horizontal error bars provide the $95^{th}$ percentile
 and illustrate the
stability of the feature importance across the leave-one-out models.
We augment the 31 feature images from the first stage by adding an additional
seven voting maps and 7 segmentation posteriors from application of the Bayesian-based
segmentation for a total of 45 images for the second stage.

__Figure 7:__ (a) FLAIR image slice illustrating WMHs which have been manually delineated.
The region around the WMHs is enlarged (b) in the original FLAIR and the
(c) contralateral FLAIR difference image.

__Figure 8:__ Voxelwise comparison with manual delineation of white matter hyperintensities.
On the left are the
calculated Dice values over all white matter hyperintensities.  Note the improvement
in the Dice metric from the employment of the Stage 2 component of the processing pipeline.
(Right) Similar results can be seen by comparing the total lesion load volume between manual
and automated detection strategies.  Although some outliers are found after the Stage 2
processing in a couple subjects, the number of outliers caused by false positives is
decreased significantly with the second stage processing.

>>>>

header-includes:
   - \usepackage{booktabs}
   - \usepackage[final]{changes}
   - \usepackage[font={small},labelfont=bf,labelsep=colon]{caption}
   - \linespread{1.2}
   - \usepackage[compact]{titlesec}
   - \usepackage{enumitem}
   - \usepackage{tikz}
   - \def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
   - \setlist{nolistsep}
   - \titlespacing{\section}{2pt}{*0}{*0}
   - \titlespacing{\subsection}{2pt}{*0}{*0}
   - \titlespacing{\subsubsection}{2pt}{*0}{*0}
   - \setlength{\parskip}{3pt}
   - \setremarkmarkup{(#2)}
   - \definechangesauthor[name={Nick Tustison}, color=red]{nt}
   - \definechangesauthor[name={James Stone}, color=magenta]{js}
   - \definechangesauthor[name={Lisa Wilde}, color=cyan]{lw}
   - \definechangesauthor[name={Andy Mayer}, color=green]{am}
   - \definechangesauthor[name={Harvey Levin}, color=blue]{hl}
   - \definechangesauthor[name={Brian Taylor}, color=purple]{bt}
   - \definechangesauthor[name={David Tate}, color=orange]{dt}
bibliography: references.bib
csl: council-of-science-editors.csl
fontsize: 11pt
mainfont: Georgia
geometry: margin=1.0in
---

@@

White matter hyperintensities (WMHs) are foci of abnormal signal
intensity in white matter regions seen with magnetic resonance imaging
(MRI). WMHs are associated with normal aging and have shown prognostic
value in neurological conditions such as traumatic brain injury (TBI).
The impracticality of manually quantifying these lesions limits their
clinical utility and motivates the utilization of machine learning
techniques for automated segmentation workflows.  Herein, we develop a
concatenated random forest framework with image features for
segmenting WMHs in a TBI cohort. The framework is built upon the
Advanced Normalization Tools (ANTs)  and ANTsR toolkits.  MR
(3D FLAIR, T2-, and T1-weighted) images from 24 service members and
veterans scanned in the Chronic Effects of Neurotrauma Consortium’s
(CENC) observational study were acquired. Manual annotations were
employed for both training and evaluation using a leave-one-out
strategy.  \textcolor{blue}{Performance measures include sensitivity,
positive predictive value, $F_1$ score, and relative volume difference.
Final average results were: sensitivity = 0.68 $\pm$ 0.38, positive
predictive value = 0.51 $\pm$ 0.40, $F_1$ = 0.52 $\pm$ 0.36,
relative volume difference} = 43 $\pm$ 26\%.  In addition, three
lesion size ranges are selected to illustrate the variation in
performance with lesion size.}  Paired with
correlative outcome data, supervised learning methods may allow for
identification of imaging features predictive of diagnosis and prognosis
in individual TBI patients.

@@

\textcolor{blue}{Figure 1 provides a graphical overview of the proposed workflow.  The major components
include offline generation of symmetric multimodal templates, the creation of
feature images from the training data which are then employed for statistical
prediction using a concatenated random forest framework.  This framework involves
the use of two random forest models where the outcome (i.e., the tissue
probability estimates) of the first RF model application,
which we denote as ``Stage1'', is used as input (along with the original set of
feature images) to a refinement RF model segmentation which we denote as ``Stage 2''.
Once these offline steps
are performed, a new (i.e., unsegmented) subject can then be processed using the proposed pipeline.}

@@

![Canonical views of the mutlivariate, bilaterally symmetric template constructed
from the MMRR data set [@landman2011] (only shown are the FLAIR, T1, and T2 modalities---
the components relevant for this work).  Template construction is detailed in
[@Tustison:2015aa].  These images are important for specific asymmetry-based
features.](Figures/MMRR.png)

@@

\textcolor{blue}{In Figure 5 we provide the segmentation evaluations derived from the leave-one-out evaluation
of the previously described TBI data over the three lesion volume ranges.
These performance measures include sensitivity,
positive predictive value, $F_1$ score, and relative volume difference.  The three lesion
size ranges over which these measures are computed are meant to illustrate the variation
in performance with lesion size.}

\textcolor{blue}{Smaller lesions ($< 12$ voxels) are more difficult
to identify which is why the sensitivity for this range is more varied compared with the largest
set of lesions ($> 28$ voxels).   The first three measures are based on the identification of
entire lesions.  The relative volume difference provides a direct assessment of the accuracy of
the volumetric estimate when comparing the manually identified lesions versus the
automatically predicted lesions.}

\textcolor{blue}{
We averaged these measures over all lesions and all subjects which resulted in the
following values:}

* \textcolor{blue}{Stage 1}

    * \textcolor{blue}{sensitivity} = 0.70 $\pm$ 0.34

    * \textcolor{blue}{PPV} = 0.42 $\pm$ 0.36

    * \textcolor{blue}{$F_1$} = 0.47 $\pm$ 0.36

    * \textcolor{blue}{relative volume difference} = 43 $\pm$ 38\%

* \textcolor{blue}{Stage 1}

    * \textcolor{blue}{sensitivity} = 0.68 $\pm$ 0.38

    * \textcolor{blue}{PPV} = 0.51 $\pm$ 0.40

    * \textcolor{blue}{$F_1$} = 0.52 $\pm$ 0.36

    * \textcolor{blue}{relative volume difference} = 43 $\pm$ 26\%

@@

In Figure 5 are the segmentation comparisons derived from manual segmentations of the same
data.  Despite the large variability characteristic with manual labelings in related fields
[@Grimaud:1996aa;@styner2008;@Garcia-Lorenzo:2013aa], such labelings are characteristic of
current clinical practices and the methodology proposed herein is readily adapted to refinements
in training data.

On the left of Figure 5 are the improvement in Dice values [@tustison2009],
over all white matter hyperintensities when comparing the segmentations between the two stages
where the sum is taken over all individually labeled manual, $T_r$, and automated, $S_r$, lesions and $\cap$
represents the intersection between the manual/automated lesion pair.  Performing
the second round of supervised learning improves these Dice values.  One can also note from the
right side of Figure 4 that the total lesion load volume illustrates a few subjects that are
severe outliers in terms of the number of false positives.  The second round helps to
correct this issue.

@@

![\textcolor{blue}{Evaluation measures for both Stages of the leave-one-out protocol of the described protocol in the Methods
section:  (a) sensitivity, (b) positive predictive value, (c) $F_1$ score, and (d) relative
volume difference.   These quantitative assessments are given for three quantile ranges
spanning the range of the manually-derived lesion volumes.    Overall improvement in all
three whole lesion-based measuers is seen as the
second Stage RF model is applied for all three quantile ranges.
The relative volume difference corresponding to the Stage 2 results tend to predict a
decreased predicted volume over the Stage 1 results.}](Figures/evaluationResults.png)

@@

The resulting rankings for both Stages are given in Figures 6 and 7 where the values for the
separate stages are averaged over the entire corresponding model set.  In addition, we
track the variance for each feature over all models to illustrate the stability of
the chosen features during the evaluation.  This latter information is illustrated as
horizontal errors bars providing the $95^{th}$ percentile
Note that the reader can cross reference Table 1 for identifying corresponding feature types
and names.

@@

\textcolor{blue}{
In evaluating the segmentation results, there are limited methodologies described in the
literature for automatic segmentation of WMHs despite the numerous clinical studies}
[@Debette:2010aa;@Kloppenborg:2014aa] \textcolor{blue}{exploring the connections between WMHs and TBI.
However, we can extrapolate from similar application areas where such methodologies have a
much longer history of development such as white matter lesion segmentation in
multiple sclerosis )MS).  In a recent review [@Garcia-Lorenzo:2013aa], 47 total papers
were found during a literature search and evaluated which concern automated segmentation of white
matter lesions in MS.  The most widely-used data for algorithmic evaluation consists
of two cohorts from two different sites used in
the MS Lesion segmentation challenge associated with the international MICCAI 2008
conference.  For comparison, this challenge training data set
has a mean lesion load of 204 ($\pm$ 752) mm$^3$ per lesion (compared with
our mean lesion load of 33.8 ($\pm$ 55.4) voxels $\times$ 1.2 mm$^3$ per voxel =
40.56 ($\pm$ 66.48) mm$^3$ per lesion) and the resolution is
almost twice what is used in this study (i.e., 0.5 $\times$ 0.5 $\times$ 0.5).
Note that our relative volume difference numbers are comparable to the
relative volume difference {\em between raters} of 68 percent (cf Figure 5 (d))}
[@vanginneken:2007aa].

Regarding the feature rankings, it is interesting to note some of the other top performing features for Stage 1.
The contralateral difference FLAIR image is  highly discriminative over the set of
evaluation random forest models (see Figure 8).  This accords with the known clinical relevance of
FLAIR images for identifying white matter hyperintensities and the fact that such
pathology does not typically manifest symmetrically in both hemispheres.  Interestingly, the
posterior maps for the deep gray matter are extremely important for accurate white matter
hyperintensity segmentation.  Perhaps the spatial specification of deep
gray matter aids in the removal of false positives.  Inspection of the bottom of the
plots demonstrates the lack of discriminating features associated with the
T1 image which is also well-known in the clinical literature.

As described earlier, for Stage 2, we used the output random forest voting maps from Stage
1 as both features themselves and as priors for input to a Bayesian-based segmentation with
an additional MRF spatial prior.  In Figure 7, the voting maps are labeled as
"_RFStage1VotingMaps_" where the final numeral is associated with the brain
parenchymal labeling given previously.  Similarly, the additional RF prior segmentation
feature probability maps are labeled as "_RFBrainSegmentationPosteriors_".
The Stage 2 feature importance plot follows similar
trends as that for Stage 1 with the T1 images not contributing much to the identification
of white matter hyperintensity voxels.  The initial voting maps from Stage 1 are extremely
important with the top 3 being the estimated locations of the 1) gray matter, 2) white matter, and
3) white matter hyperintensities.  Since these tissue type can be conflated based on intensity
alone it is intuitive that such features would be important.

@@
